Bayesian Rate Estimation

### 2012-07-31 many.tex
Major edits.
* Updated the derivation of the likelihood, approaching first from the
  normalized foreground and background distributions, and treating the
  Poisson factor for the rates as a sort of ``prior'' for the rate
  flags.

* Added two examples:

  * Fitting the effective number of independent templates from the a
    Gaussian-noise data set as a shape parameter of the background
    distribution.

  * An example of fitting globular cluster parameters from observed
    stars given a non-uniform background.

\documentclass[aps,prd,reprint]{revtex4-1}

@@

\newcommand{\order}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\fixme}[1]{\textbf{FIXME: #1}}

@@

\author{Will M. Farr}
\email{w-farr@northwestern.edu}
%\homepage{http://faculty.wcas.northwestern.edu/will-farr/}
% \affiliation{Center for Interdisciplinary Exploration and Research in Astrophysics (CIERA)\\
% Department of Physics and Astronomy\\
% Northwestern University, 2145 Sheridan Road, Evanston, IL 60208}
\affiliation{Northwestern University and CIERA}

\author{Ilya Mandel}
\email{ilya@chgk.info}
%\homepage{http://www.chgk.info/~ilyamandel/}
% \affiliation{School of Physics and Astronomy\\University of Birmingham\\Edgbaston B15 2TT Birmingham\\United Kingdom}
\affiliation{University of Birmingham}

\author{Jon Gair}

@@

We assume that we are presented with a data set of $N$ events.  Each
event may be due to either a signal of interest or an uninteresting
background.  Each event is associated with a ranking statistic, $x$.
Our data set therefore consists of the ranking statistics for the set
of events:
\begin{equation}
  d = \{ x_i | i = 1, \ldots, N \}.
\end{equation}

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process with rates (per unit ranking
statistic, $x$)
\begin{equation}
  \frac{dN_f}{dx} = f(x)
\end{equation}
and 
\begin{equation}
  \frac{dN_b}{dx} = b(x).
\end{equation}
The cumulative rates of the two processes are therefore
\begin{equation}
  F(x) \equiv \int_{-\infty}^x ds\, f(s)
\end{equation}
and
\begin{equation}
  B(x) \equiv \int_{-\infty}^x ds\, b(s).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item The number of events in any range of ranking statistics, $x \in
  [x_1, x_2]$ is Poisson distributed with rate $F(x_2) - F(x_1)$ or
  $B(x_2) - B(x_1)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent. 
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(N = 1 \in [x, x+h]) = f(x) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligable
  \begin{equation}
    P(N = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x) = R_f \hat{f}(x),
\end{equation}
and 
\begin{equation}
  b(x) = R_b \hat{b}(x),
\end{equation}
where $\hat{F}(\infty) = \hat{B}(\infty) = 1$.  Then $R_f$ and $R_b$
are the total number of foreground and background events expected,
respectively.  Other parameters may describe the shape of the rate
functions, but these will depend on the details of the system being
analyzed.  In \S \ref{sec:GW-example} we give an example of fitting
such shape paremeters.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 or 1, indicating a background or foreground event,
respectively.  These ``state'' flags are additional (unobserved) data
in our model.  We can marginalize over our uncertainty in the state of
any given event by summing all posteriors over $f_i = \{0,1\}$.  Given
an identification of each event as foreground or background, and the
rates of each, the likelihood%
\footnote{See \S \ref{sec:likelihood-derivation} for a derivation of
  Eq.~\eqref{eq:likelihood}.} %
of our data is
\begin{multline}
  \label{eq:likelihood}
  p\left( d, \left\{ f_i \right\} | f(x), b(x)
  \right) = \\ \left[ \prod_{\left\{ i | f_i = 1 \right\}} f\left( x_i
    \right) \right]  \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    b\left( x_j \right) \right] \\
  \times \exp\left[ - F\left( \infty \right) \right] 
  \exp\left[ -B\left( \infty \right) \right].
\end{multline}
Written in terms of the rate parameters, this becomes
\begin{multline}
  \label{eq:likelihood-rates}
  p\left( d, \left\{ f_i \right\} | f(x), b(x)\right) = \\ R_f^{N_f}
  \left[ \prod_{\left\{ i | f_i = 1 \right\}}
    \hat{f}\left( x_i \right) \right] \exp\left[ - R_f \right] \\
  \times R_b^{N_b} \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    \hat{b}\left( x_j \right) \right] \exp\left[ - R_b \right],
\end{multline}
where $N_f$ is the number of the $f_i$ that are 1 (i.e.\ the number of
assumed foreground events), and $N_b$ is the number of the $f_i$ that
are 0 (i.e.\ the assumed number of background events).  Note that $N_f
+ N_b = N$, as each event is considered either foreground or
background in our model.

It will occasionally be convenient to work with the ratio,
$\hat{f}(x)/\hat{b}(x)$ instead of the values of $\hat{f}$ and
$\hat{b}$ separately.  This ratio can be interpreted as a likelihood
ratio between the foreground and background models for an event that
has occurred with ranking statistic $x$:
\begin{equation}
  \label{eq:loudness-likelihood}
  \frac{\hat{f}(x)}{\hat{b}(x)} = \frac{p(x|\mathrm{foreground})}{p(x|\mathrm{background})}.
\end{equation}
In some circumstances, it is possible to compute this ratio
empirically (for example, in \citet{Cannon2012}, the authors give a
practical method for estimating the distribution of loudnesses for
known-background events and known-foreground events in a mock data
stream from the Advanced LIGO and Advanced Virgo gravitational wave
detectors).  Dividing Eq.~\eqref{eq:likelihood-rates} by
$\prod_{\left\{ i | f_i = 0 \right\}} \hat{b}(x_i)$, we obtain
\begin{multline}
  \label{eq:likelihood-ratio}
  p\left(d, \left\{ f_i \right\} | f(x), b(x) \right) \propto  \\ \left[
    \prod_{\left\{ i | f_i = 1 \right\}}
    \frac{\hat{f}(x_i)}{\hat{b}(x_i)} \right] R_f^{N_f} R_b^{N_b}
  \exp\left[ - \left( R_f + R_b \right) \right].
\end{multline}

\subsection{Priors}

Because the overall rates enter the likelihood in
Eq.~\eqref{eq:likelihood-rates} in the same form as in the
constant-rate Poisson likelihood, a reasonable prior for $R_f$ and
$R_b$ is the Poisson Jeffrey's prior%
\footnote{Recall that the Jeffrey's prior for a parameter $\theta$
  with likeilhood function, $p(d|\theta)$, is
  \begin{equation}
    p(\theta) \propto \sqrt{\left\langle \left( \frac{\partial
          p(d | \theta) }{\partial \theta} \right)^2 \right\rangle},
  \end{equation}
  where the expectation is taken over all data, $d$.  The Jeffrey's
  prior assigns prior weight to locations in parameter space
  proportional to the Fisher information that is expected to result
  from a measurement at that parameter value.}, %
\begin{equation}
  \label{eq:Jeffrey-prior}
  p(R) \propto \frac{1}{\sqrt{R}}.
\end{equation}
One advantage of the Jeffrey's prior in this circumstance is that it
has finite probability mass as $R \to 0$, and the likelihood in
Eq.~\eqref{eq:likelihood-rates} controls its behavior as $R\to \infty$
to ensure a normalizable posterior.

Of course, other priors could be used; in particular, previous
experiments may have placed constraints on the rate that can be used
as priors in subsequent experiments.  Throughout the remainder of this
paper, however, we use the Jeffrey's prior,
Eq.~\eqref{eq:Jeffrey-prior}.

\subsection{Posteriors}

Putting everything together, the posterior probability of each event's
state and the overall rate is given by
\begin{multline}
  p(R_f, R_b, \left\{ f_i \right\} | d) \propto R_f^{N_f}
  \left[ \prod_{\left\{ i | f_i = 1 \right\}}
    \hat{f}\left( x_i \right) \right] \exp\left[ - R_f \right] \\
  \times R_b^{N_b} \left[ \prod_{\left\{ j | f_j = 0 \right\}}
    \hat{b}\left( x_j \right) \right] \exp\left[ - R_b \right] p(R_f, R_b).
\end{multline}
If we marginalize over all the flags, $f_i$, then we obtain
\begin{multline}
  \label{eq:rates-marginalized}
  p(R_f, R_b | d) \propto \prod_{i} \left[ R_f \hat{f}\left( x_i
    \right) + R_b \hat{b}\left( x_i \right) \right] \\ \times
  \exp\left[ - \left( R_f + R_b \right) \right] p\left( R_f, R_b
  \right).
\end{multline}
If, instead, we marginalize over all but one flag, $f_k$, we obtain
that the posterior ratio of signal to noise probability for event $k$
given the rates $R_f$ and $R_b$ is 
\begin{equation}
  \label{eq:foreground-marginalized}
  \frac{p(R_f, R_b, f_k = 1 | d)}{p(R_f, R_b, f_k = 0 | d)} =
  \frac{R_f \hat{f}\left( x_k \right)}{R_b \hat{b}\left( x_k
    \right)}p\left(R_f, R_b\right). 
\end{equation}
Note that the \emph{posterior} estimate of the ratio of signal to
noise probabilities for event $i$ differs from the likelihood ratio,
Eq.~\eqref{eq:loudness-likelihood}, because it depends on the overall
rates of foreground and background events.

\section{Gravitational Waves from Compact Binary Inspirals}
\label{sec:GW-example}

\subsection{Analytic Example}
\label{sec:analytic-GW-example}

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering against a set of $N$ templates.  The data
stream consists of stationary Gaussian noise with a power spectral
density $S(f)$ combined additively with some number of signals.  The
signal to noise ratio (SNR) of a template, $h(f)$, given data, $d(f)$,
is given by
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
        \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that 
\begin{equation}
  \left\langle h_1, h_2 \right\rangle \sim 0.
\end{equation}

For a data stream of pure noise, $d(f) = n(f)$, the SNR of a each
template follows a $N(0,1)$ distribution.  If our ranking statistic
for events in each interval of data is their maximum SNR over the $N$
templates, 
\begin{equation}
  x \equiv \max_h \rho_h,
\end{equation}
then the background distribution is
\begin{equation}
  \hat{B}(x) = \left( \frac{1 + \erf\left( \frac{x}{2}
      \right)}{2} \right)^N
\end{equation}
If, in addition, we impose a threshold ranking statistic, $\xmin$, to
be considered an ``event,'' then $\hat{B}$ becomes
\begin{equation}
  \label{eq:analytic-background-rate}
  \hat{B}(x) = \frac{\left( 1 + \erf\left( \frac{x}{2} \right)
    \right)^N - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^N}{2^N - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^N }
\end{equation}

The SNR of a gravitational wave signal in an interferometric detector
scales as $1/d$, where $d$ is the distance to the source.  Ignoring
possible cosmological effects, the number of sources will scale as
$d^3$.  Thus, we expect that the foreground cumulative distribution of
events will follow
\begin{equation}
  \label{eq:analytic-foreground-rate}
  \hat{F}(x) = 1 - \frac{\xmin^3}{x^3}.
\end{equation}

To demonstrate the effectiveness of our formalism, we applied it to a
synthetic data set with foreground and background distributions drawn
from Eqs.~\eqref{eq:analytic-background-rate} and
\eqref{eq:analytic-foreground-rate} with $R_f^\mathrm{true} = 10.4$
and $R_b^\mathrm{true} = 95.1$.  The cumulative distribution for the
ranking statistic of the synthetic data we used appears in Figure
\ref{fig:analytic-data-cumulative}.

@@

In Figure \ref{fig:analytic-rate-recovery}, we show the marginalized
posterior densities for the foreground and background rates.  (Refer
to Eq.~\eqref{eq:rates-marginalized}.)  Figure
\ref{fig:analytic-rate-foreground-probs} shows the posterior
foreground probability for each event marginalized over all other
events' types and the foreground and background rates.  (Refer to
Eq.~\eqref{eq:foreground-marginalized}.)

\begin{figure}
  \includegraphics[width=\columnwidth]{rates}
  \caption{\label{fig:analytic-rate-recovery} The posterior densities
    for $R_f$ (in blue) and $R_b$ (in black) for the analytic model
    discussed in \S \ref{sec:analytic-GW-example}.  The vertical lines
    indicate the ``true'' used to generate the synthetic data set.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters.  True foreground events are in blue, background events
    in red.  The solid line is the likelihood ratio
    $p(x|\mathrm{foreground})/p(x|\mathrm{background})$ (see
    Eq.~\eqref{eq:loudness-likelihood}) for this model; this exceeds
    the marginalized foreground probability for many of the events
    because in this data set there are approximately nine times as
    many background events as foreground events (see
    Eq.~\eqref{eq:foreground-marginalized}). }
\end{figure}

\begin{acknowledgments}
  Kipp Cannon, Chad Hanna, and Drew Keppel for providing the data set
  from \citet{Cannon2012} and assisting us in its use. Richard
  O'Shaughnessy for discussions.
\end{acknowledgments}

\appendix

\section{Likelihood for Inhomogeneous Poisson Processes}
\label{sec:likelihood-derivation}

\fixme{To do.}

>>>>

\documentclass[aps,prd]{revtex4-1}

@@

\newcommand{\order}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\fixme}[1]{\textbf{FIXME: #1}}
\newcommand{\mathset}[1]{\left\{ #1 \right\}}

@@

\author{Will M. Farr}
\email{w-farr@northwestern.edu}
\homepage{http://faculty.wcas.northwestern.edu/will-farr/}
\affiliation{Center for Interdisciplinary Exploration and Research in Astrophysics\\
Department of Physics and Astronomy\\
Northwestern University, 2145 Sheridan Road, Evanston, IL 60208}
%\affiliation{Northwestern University and CIERA}

\author{Ilya Mandel}
\email{ilya@chgk.info}
\homepage{http://www.chgk.info/~ilyamandel/}
\affiliation{School of Physics and Astronomy\\University of Birmingham\\Edgbaston B15 2TT Birmingham\\United Kingdom}
%\affiliation{University of Birmingham}

\author{Jon Gair}
\email{jrg23@cam.ac.uk}

@@

We assume that we are presented with a data set of $N$ events.  Each
event may be due to either a signal of interest or an uninteresting
background.  Each event is associated with a ranking statistic, $x$.
Our data set therefore consists of the ranking statistics for the set
of events:
\begin{equation}
  d = \mathset{ x_i | i = 1, \ldots, N }
\end{equation}

We assume that both the foreground and background events are samples
from an inhomogeneous Poisson process with differential rates
\begin{equation}
  \frac{dN_f}{dx} = f(x, \theta)
\end{equation}
and 
\begin{equation}
  \frac{dN_b}{dx} = b(x, \theta),
\end{equation}
where the $\theta$ argument represents additional ``shape'' parameters
that may affect the distribution, and for which we will eventually
fit.  The cumulative rates of the two processes are therefore
\begin{equation}
  F(x,\theta) \equiv \int_{-\infty}^x ds\, f(s, \theta)
\end{equation}
and
\begin{equation}
  B(x,\theta) \equiv \int_{-\infty}^x ds\, b(s,\theta).
\end{equation}
The assumption that the foreground and background events form an
inhomogeneous Poisson process implies
\begin{enumerate}
\item \label{prop:Poisson} The number of events in any range of
  ranking statistics, $x \in [x_1, x_2]$ is Poisson distributed with
  rate $F(x_2, \theta) - F(x_1,\theta)$ or $B(x_2,\theta) -
  B(x_1,\theta)$.
\item The numbers of events in non-overlapping ranges of ranking
  statistics are independent. 
\item The probability of exactly one foreground event between $x$ and
  $x+h$ is given by
  \begin{equation}
    P(N = 1 \in [x, x+h]) = f(x,\theta) h + \order{h^2}.
  \end{equation}
  and similarly for background events.
\item The probability of two or more events in a small range of
  ranking statistic is negligable
  \begin{equation}
    P(N = 2 \in [x, x+h]) = \order{h^2}.
  \end{equation}
\end{enumerate}
The foreground and background rates can in general depend on several
parameters; the goal of our analysis is to determine the posterior
probability distributions for these parameters that are implied by the
data.  At the least, we will want to know the overall amplitude of the
foreground and background rates.  Let
\begin{equation}
  f(x,\theta) = R_f \hat{f}(x,\theta'),
\end{equation}
and 
\begin{equation}
  b(x, \theta) = R_b \hat{b}(x, \theta'),
\end{equation}
where $\hat{F}(\infty, \theta') = \hat{B}(\infty, \theta') = 1$, and
$\theta' = \theta \setminus \{R_{f}, R_{b} \}$.  Then $R_f$ and $R_b$
are the total number of foreground and background events expected and
$\hat{f}(x, \theta')$ and $\hat{b}(x, \theta')$ are the likelihood of
obtaining an event with ranking statistic $x$ under the foreground and
background distributions.  In what follows, we will drop the prime,
using $\theta$ to denote all parameters of the rate distributions
except $R_f$ and $R_b$.

We do not know a priori which of the events are foreground and which
are background.  For each event, we introduce a flag, $f_i$, which is
either 0 or 1, indicating a background or foreground event.  These
``state'' flags are parameters in our model, along with $R_f$, $R_b$,
and $\theta$.  We can marginalize over our uncertainty in the state of
any given event by summing posteriors over $f_i = \mathset{0,1}$.

Bayes' theorem relates the posterior probability of the state flags,
rates, and shape parameters, $p\left(\mathset{f_i}, R_f, R_b, \theta |
  d \right)$, the likelihood of the data, $p\left( d | \mathset{f_i},
  R_f, R_b, \theta\right)$, and the prior probability of state flags,
rates and shape parameters before any data are obtained, $p\left(
  \mathset{f_i}, R_f, R_b, \theta\right)$:
\begin{equation}
  \label{eq:Bayes}
  p\left(\mathset{f_i}, R_f, R_b, \theta | d \right)  \propto
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)
  p\left(\mathset{f_i}, R_f, R_b, \theta\right). 
\end{equation}
The constant of proportionality, called the evidence, $p(d)$, is a
normalization constant independent of the state flags, rates, and
shape parameters.

Each foreground event is drawn from the probability distribution
$\hat{f}$ and each background event is drawn from the probability
distribution $\hat{b}$.  The events are independent of each other.
Therefore, the likelihood of the data is 
\begin{equation}
  p\left( d | \mathset{f_i}, R_f, R_b, \theta\right)  = \left[
    \prod_{\mathset{i | f_i = 1}} \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} \hat{b}\left( x_i, \theta\right)
  \right].  
\end{equation}
The prior distribution can be factorized as
\begin{equation}
  \label{eq:combined-flag-rate-prior}
  p\left(\mathset{f_i}, R_f, R_b, \theta\right)  = p\left(
    \mathset{f_i} | R_f, R_b, \theta\right)p(R_f, R_b, \theta).
\end{equation}
The first term is not acutally a ``prior'' in the usual sense; the
conditional probability of the $\mathset{f_i}$ on the rates is given
by
\begin{equation}
  \label{eq:flag-conditional-prior}
  p\left(\mathset{f_i} | R_f, R_b, \theta\right) = \frac{R_f^{N_f}
    R_b^{N_b}}{N_f! N_b!} \exp\left[ - \left(R_f + R_b\right) \right]
  N_f! N_b! = R_f^{N_f}R_b^{N_b} \exp\left[ - \left(R_f + R_b\right) \right],
\end{equation}
where $N_f$ and $N_b$ are the number of foreground and background
events.  This expression follows from Property \ref{prop:Poisson} of
inhomogenous Poisson processes and the $N_f! N_b!$ differnt ways of
assigning foreground and background flags to events for fixed $N_f$,
$N_b$.

The second term in Eq.~\eqref{eq:combined-flag-rate-prior} is a
traditional prior.  Because the rate parameters enter the posterior in
the same form as Poisson rates, we choose here the Poisson Jeffrey's
prior on rates, independent of the shape parameters
\begin{equation}
  p\left( R_f, R_b, \theta\right) = \frac{1}{\sqrt{R_f R_b}} p(\theta),
\end{equation}
but of course other choices are possible.  This choice has the
advantage that the prior is normalizable as $R_f, R_b \to 0$, and the
exponentials in Eq.~\eqref{eq:flag-conditional-prior} regularize the
posterior as $R_f, R_b \to \infty$.

Putting everything together, the posterior is
\begin{equation}
  \label{eq:posterior}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto \left[
    \prod_{\mathset{i | f_i = 1}} R_f \hat{f}\left(x_i, \theta\right) \right]
  \left[ \prod_{\mathset{i | f_i = 0}} R_b \hat{b}\left( x_i, \theta\right)
  \right] \exp\left[ - \left(R_f + R_b\right) \right]
  \frac{p(\theta)}{\sqrt{R_f R_b}}
\end{equation}
Not surprisingly, this posterior equivalent to one derived by assuming
that the flags, $\mathset{f_i}$, are un-observed data and treating the
sets $\mathset{x_i | f_i = 1}$ and $\mathset{x_i | f_i = 0}$ as
samples from an inhomogeneous Poisson process:
\begin{multline}
  p\left( \mathset{f_i}, R_f, R_b, \theta | d \right) \propto p\left(
    d, \mathset{f_i} | R_f, R_b, \theta \right) p\left( R_f, R_b,
    \theta \right) \\ = \left[ \prod_{\mathset{i|f_i=1}} f\left( x_i,
      \theta \right) \right] \left[ \prod_{\mathset{i|f_i=0}} b\left(
      x_i, \theta\right) \right] \exp\left[ -\left( R_f + R_b\right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{multline}

We can marginalize the posterior over the flags, $f_i$, obtaining
\begin{equation}
  \label{eq:rate-shape-posterior}
  p\left( R_f, R_b, \theta | d \right) = \sum_{\mathset{f_i} \in
    \mathset{0,1}^N} p\left( \mathset{f_i}, R_f, R_b, \theta | d \right)
  \propto \prod_{i} \left[ R_f \hat{f}\left(x_i, \theta\right) + R_b
    \hat{b}\left( x_i, \theta\right) \right] \exp\left[-\left( R_f +
      R_b \right) \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
This expression is useful if we are only interested in rates and not
the probability that any particular event is foreground or background.
Unlike the full posterior (Eq.~\eqref{eq:posterior}),
Eq.~\eqref{eq:rate-shape-posterior} contains only continuous
parameters.

Eq.~\eqref{eq:posterior} is unchanged if the ranking statistic is
multi-dimensional; in this case, the rates are 
\begin{equation}
  R_f = \int d^n x \, f(x, \theta)
\end{equation}
and
\begin{equation}
  R_b = \int d^n x \, b(x, \theta),  
\end{equation}
where $f$ and $b$ are rate densities on the $n$-dimensional space of
ranking statistics.  We give an example of fitting for
multi-dimensional rate densities in \S~\ref{sec:star-cluster}.

\section{Examples}
\label{sec:GW-example}

In this section we present several examples of the application of our
framework to various rate estimation problems in the presence of
background.

\subsection{Gravitational Waves with Non-Overlapping Templates}
\label{sec:analytic-GW-example}

Suppose we attempt to detect gravitational wave signals in a data
stream by matched filtering in the frequency domain against a set of
$N$ template waveforms \citep[e.g.][]{LVC2011}.  In our simplistic
model, we suppose the data stream consists of stationary Gaussian
noise with a power spectral density $S(f)$ combined additively with
some number of gravitational wave signals.  We assume that the signals
are sufficiently rare that they do not overlap in the data stream.
The signal to noise ratio (SNR) of a template, $h(f)$, given data,
$d(f)$, is
\begin{equation}
  \rho_h \equiv \frac{\left\langle h, d \right\rangle}{\sqrt{\left
        \langle h, h \right\rangle}},
\end{equation}
where $\left \langle \cdot \right\rangle$ denotes the noise-weighted
inner product:
\begin{equation}
  \left\langle a, b \right\rangle \equiv \int_0^\infty df\,
  \frac{a^*(f) b(f)}{S(f)}.
\end{equation}
We suppose for simplicity that the templates are sufficiently distinct
that 
\begin{equation}
  \left\langle h_i, h_j \right\rangle \simeq \delta_{ij}.
\end{equation}
In the following subsection, we will generalize the model to
overlapping templates.  We rank candidate events by their maxmim SNR
over the entire template bank,
\begin{equation}
  x \equiv \max_{h} \rho_h,
\end{equation}
and consider only events that have a maximum SNR above some threshold,
$x > \xmin$.

For a data stream of pure noise, $d(f) = n(f)$, the SNR of a each
template follows a $N(0,1)$ distribution.  The background ranking
statistic (i.e.\ the maximum SNR over the template bank) then has a
cumulative distribution without thresholding of
\begin{equation}
  \hat{B}(x) = \left( \frac{1 + \erf\left( \frac{x}{2}
      \right)}{2} \right)^N
\end{equation}
Imposing the threshold, $x > \xmin$, the
cumulative distribution of the background becomes
\begin{equation}
  \label{eq:analytic-background-rate}
  \hat{B}(x) = \frac{\left( 1 + \erf\left( \frac{x}{2} \right)
    \right)^N - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^N}{2^N - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^N }
\end{equation}

The SNR of a gravitational wave signal in an interferometric detector
scales as $1/d$ \citep{Finn1992}, where $d$ is the distance to the
source.  Ignoring cosmological effects, the number of sources scales
as $d^3$.  Thus, we expect that the foreground cumulative distribution
of events will follow
\begin{equation}
  \label{eq:analytic-foreground-rate}
  \hat{F}(x) = 1 - \frac{\xmin^3}{x^3}.
\end{equation}

Note that this scenario has no shape parameters for the foreground and
background distributions.

To demonstrate the effectiveness of our formalism, we applied it to a
synthetic data set with foreground and background distributions drawn
from Eqs.~\eqref{eq:analytic-background-rate} and
\eqref{eq:analytic-foreground-rate} with $R_f^\mathrm{true} = 10.4$
and $R_b^\mathrm{true} = 95.1$, using 1000 templates.  The synthetic
data consisted of 13 foreground events and 85 background events; the
cumulative distribution for the ranking statistic of the synthetic
data appears in Figure \ref{fig:analytic-data-cumulative}.  We used a
Markov Chain Monte Carlo simulation to draw samples of state flags and
rates from the joint posterior (Eq.~\eqref{eq:posterior}).

@@

In Figure \ref{fig:analytic-rate-recovery}, we show the marginalized
posterior densities for the foreground and background rates.  (Refer
to Eq.~\eqref{eq:rate-shape-posterior}.)  Figure
\ref{fig:analytic-rate-foreground-probs} shows the posterior
foreground probability for each event marginalized over all other
events' types and the foreground and background rates.

\begin{figure}
  \includegraphics[width=\columnwidth]{rates}
  \caption{\label{fig:analytic-rate-recovery} The marginalized
    posterior densities for $R_f$ (solid line) and $R_b$ (dashed line)
    for the analytic model discussed in \S
    \ref{sec:analytic-GW-example}.  The vertical lines indicate the
    ``true'' values used to generate the synthetic data set.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters.  True foreground events are in blue, background events
    in red.  The solid line is the likelihood ratio
    $p(x|\mathrm{foreground})/p(x|\mathrm{background}) =
    \hat{f}(x)/\hat{b}(x)$ for this model; this exceeds the
    marginalized foreground probability for many of the events because
    in this data set there are approximately nine times as many
    background events as foreground events. }
\end{figure}

\subsection{Gravitational Waves With Overlapping Templates}
\label{sec:gw-overlapping-template}

In \S \ref{sec:analytic-GW-example} we assumed that the overlap
between templates in the template bank vanished, so the SNR's in
different templates are independent random variables.  In fact,
template banks are not constructed in this way
\citep{Caudill2011,Cannon2011}, because signals can fall in the gaps
between the non-overlapping templates.  We can model this effect by
assuming that a template bank of $N$ actual templates will behave as
if it had $N_\mathrm{eff}$ \emph{independent} templates, and fit for
the shape parameter $N_\mathrm{eff}$.  That is, we assume that
$N_\mathrm{eff}$ is a shape parameter for the background cumulative
distribution:
\begin{equation}
  \hat{B}\left(x, N_\mathrm{eff}\right) = \frac{\left( 1 + \erf\left( \frac{x}{2} \right)
    \right)^{N_\mathrm{eff}} - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^{N_\mathrm{eff}}}{2^{N_\mathrm{eff}} - \left( 1 + \erf\left( \frac{\xmin}{2} \right)
    \right)^{N_\mathrm{eff}} }.
\end{equation}

Results from such an analysis appear in Figures \ref{fig:rates-nt} and
\ref{fig:ntemplates}.  We use the same parameters and data set as in
\S \ref{sec:analytic-GW-example}, with $R_f = 10.4$, $R_b = 95.1$, and
$N_\mathrm{eff} = 1000$, but now allow $N_\mathrm{eff}$ to be a
parameter of the background distribution, with a flat prior.  Both the
rates and the number of effective templates are recovered without
significant loss of accuracy relative to the fixed $N_\mathrm{eff}$
situation in \S \ref{sec:analytic-GW-example}.

\begin{figure}
  \includegraphics[width=\columnwidth]{rates-nt}
  \caption{\label{fig:rates-nt} The foreground (solid lines) and
    background (dashed lines) rate posterior, marginalized over all
    flags and the $N_\mathrm{eff}$ parameter, for the gravitational
    wave template detection scenario with overlapping templates
    discussed in \S \ref{sec:gw-overlapping-template}.  The true
    values of the rates, $R_f = 10.4$ and $R_b=95.1$, are indicated
    with vertical lines.  The distributions are not significantly
    wider than those of Figure \ref{fig:analytic-rate-recovery}, in
    spite of the extra parameter.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{ntemplates}
  \caption{\label{fig:ntemplates} The posterior on the number of
    effective templates, $N_\mathrm{eff}$, for the model and data
    discussed in \S \ref{sec:gw-overlapping-template}, marginalized
    over all state flags and rates.  The true value, $N_\mathrm{eff} =
    1000$, is indicated by the vertical line.}
\end{figure}

\subsection{Star Cluster Parameters With Background Contamination}
\label{sec:star-cluster}

Our final example concerns fitting for the location and shape
parameters of a cluster of stars observed on top of a stellar
background with a density gradient.  We assume that a star cluster has
a Plummer surface-density profile \citep{Plummer1911,Aarseth1974},
\begin{equation}
  \label{eq:plummer-surface-density}
  \hat{f}(\vec{x}, \theta) = \frac{1}{\pi r_0^2 \left( 1 + \frac{\left|
          \vec{x} - \vec{x}_0 \right|^2}{r_0^2} \right)^2},
\end{equation}
where $\vec{x}_0$ is the location on the sky of the center of the
cluster, and $r_0$ is a radial scale parameter.  We assume a square
observational domain\footnote{The observational domain is not
  infinite, so the normalization of the cluster density in
  Eq.~\eqref{eq:plummer-surface-density} is not quite correct.  In our
  modeling we properly take this into account, but for simplicity here
  we ignore it.}, $\vec{x} \in [0,1]^2$, and a background that has a
density gradient at an arbitrary orientation with respect to the
observational axes:
\begin{equation}
  \hat{b}\left(\vec{x}, \theta\right) = 1 + \vec{\gamma} \cdot \left( \vec{x}
    - \vec{x}_{1/2} \right),
\end{equation}
where $\vec{\gamma}$ is the gradient, and $\vec{x}_{1/2} = [1/2, 1/2]$
is the centroid of the observational domain.  We use assume a gradient
for the background of $\vec{\gamma} = [-0.5, 0.5]$.  We assume that
the total number of background stars, $R_b = 10000$, exceeds by a
factor of ten the total number of cluster stars, $R_f = 1000$.  We
choose a true cluster center $\vec{x}_0 = \vec{x}_{1/2}$, and a radial
scale parameter of $r_0 = 0.1$.  In all, the shape parameters for the
true distributions are
\begin{equation}
\label{eq:true-cluster-parameters}
\theta_0 \equiv \left\{ x_0, y_0, r_0, \gamma_x, \gamma_y \right\} =
\left\{ \frac{1}{2}, \frac{1}{2}, \frac{1}{10}, -\frac{1}{2},
  \frac{1}{2} \right\}.
\end{equation}
Figure \ref{fig:sky-density} shows the density on the sky due to the
cluster foreground and the background with these parameters.  We drew
a synthetic data set from this density according to the corresponding
inhomogeneous Poisson process.

\begin{figure}
  \includegraphics[width=\columnwidth]{sky-density}
  \caption{\label{fig:sky-density} The density contours on the sky of
    the forergound cluster and background discussed in \S
    \ref{sec:star-cluster}, with true parameters given in
    Eq.~\eqref{eq:true-cluster-parameters}.  There are a factor of 10
    more stars (10000) in the background than in the cluster (1000),
    but the peak density for the cluster is much higher.}
\end{figure}
